Descriptive Complexity for Counting Complexity Classes

### 2018-12-09 LMCS/syntactic-clo.tex
all direct replacements by reviewer 1

It was proved in \cite{SalujaST95} that every function in $\E{1}$ admits an FPRAS. Besides, it can be proved that $\E{1} \subseteq \totp$. 
However, this class is not closed under sum, and then it is not robust under basic closure properties. 
\begin{prop}\label{prop-e1-nc}
There exist functions $f, g \in \E{1}$ such that $(f + g) \not\in \E{1}$.
\end{prop}
\proof
\input{proof-prop-e1-nc}
\qed
To overcome this limitation, one can consider the class $\QE{1}$, which is closed under sum by definition. In fact, the following proposition shows that the same good properties as for $\E{1}$ hold for $\QE{1}$, together with the fact that it is closed under sum and multiplication.
\begin{prop} \label{prop:qe0-fp-qe1-totp-fptras}
$\QE{1} \subseteq \totp$ and every function in $\QE{1}$ has an FPRAS. Moreover, $\QE{1}$ is closed under sum and multiplication.
\end{prop}
\proof
\input{proof-qe0-fp-qe1-totp-fptras}
\qed
Hence, it only remains to prove that $\QE{1}$ is closed under subtraction by one. Unfortunately, it is not clear whether this property holds; in fact, we conjecture that it is not the case. Thus, we need to find an extension of $\QE{1}$ that keeps all the previous properties and is closed under subtraction by one. It is important to notice that $\shp$ is believed not to be closed under subtraction by one by some complexity-theoretical assumption\footnote{A decision problem $L$ is in the randomized complexity class $\cspp$ if there exists a polynomial-time NTM $M$ such that for every $x \in L$ it holds that $\tma_M(x) - \tmr_M(x) = 2$, and for every $x \not\in L$ it holds that $\tma_M(x) = \tmr_M(x)$ \cite{OH93,FFK94}. It is believed that $\np \not\subseteq \cspp$.
However, if $\shp$ is closed under subtraction by one, then it holds that $\np \subseteq \cspp$ \cite{OH93}.}. So, the following proposition rules out any logic that extends $\Pi_1$ for a possible extension of~$\QE{1}$ with the desired closure property.
\begin{prop} \label{pi-minusone}
If $\Pi_1 \subseteq \LL \subseteq \fo$ and $\eqso(\LL)$ is closed under subtraction by one, then $\shp$ is closed under subtraction by~one. 
\end{prop}
\proof
\input{proof-pi-minusone}
\qed
Therefore, the desired extension has to be achieved by allowing some local extensions to~$\Sigma_1$. More precisely, we define $\logex{1}$ as $\Sigma_1$ but allowing atomic formulae over a signature~$\R$ to be of the form either $u = v$ or $X(\bar u)$, where $X$ is a second-order variable, or $\varphi(\bar u)$, where $\varphi(\bar u)$ is a first-order formula over $\R$ (in particular, it does not mentioned any second-order variable). With this extension we obtain a class with the desired properties.
\begin{thm}\label{sigmafo-minusone}
The class $\eqso(\logex{1})$ is closed under sum, multiplication and subtraction by one. Moreover, $\eqso(\logex{1}) \subseteq \totp$ and every function in $\eqso(\logex{1})$ has an FPRAS.
\end{thm}
\proof
\input{proof-sigmafo-minusone}
\qed
The proof that $\eqso(\logex{1})$ is closed under subtraction by one is the most involved of the paper. We think the main technique used in this proof, which is based on considering some witnesses of logarithmic size, is of independent interest.

>>>>

It was proved in \cite{SalujaST95} that every function in $\E{1}$ admits an FPRAS. Besides, it can be proved that $\E{1} \subseteq \totp$. 
However, this class is not closed under sum, and then it is not robust under basic closure properties. 
\begin{prop}\label{prop-e1-nc}
There exist functions $f, g \in \E{1}$ such that $(f + g) \not\in \E{1}$.
\end{prop}
\proof
\input{proof-prop-e1-nc}
\qed
To overcome this limitation, one can consider the class $\QE{1}$, which is closed under sum by definition. In fact, the following proposition shows that the same good properties as for $\E{1}$ hold for $\QE{1}$, together with the fact that it is closed under sum and multiplication.
\begin{prop} \label{prop:qe0-fp-qe1-totp-fptras}
$\QE{1} \subseteq \totp$ and every function in $\QE{1}$ has an FPRAS. Moreover, $\QE{1}$ is closed under sum and multiplication.
\end{prop}
\proof
\input{proof-qe0-fp-qe1-totp-fptras}
\qed
Hence, it only remains to prove that $\QE{1}$ is closed under subtraction by one. Unfortunately, it is not clear whether this property holds; in fact, we conjecture that it is not the case. Thus, we need to find an extension of $\QE{1}$ that keeps all the previous properties and is closed under subtraction by one. It is important to notice that $\shp$ is believed not to be closed under subtraction by one by some complexity-theoretical assumption\footnote{A decision problem $L$ is in the randomized complexity class $\cspp$ if there exists a polynomial-time NTM $M$ such that for every $x \in L$ it holds that $\tma_M(x) - \tmr_M(x) = 2$, and for every $x \not\in L$ it holds that $\tma_M(x) = \tmr_M(x)$ \cite{OH93,FFK94}. It is believed that $\np \not\subseteq \cspp$.
However, if $\shp$ is closed under subtraction by one, then it holds that $\np \subseteq \cspp$ \cite{OH93}.}. So, the following proposition rules out any logic that extends $\Pi_1$ as a possible extension of~$\QE{1}$ with the desired closure property.
\begin{prop} \label{pi-minusone}
If $\Pi_1 \subseteq \LL \subseteq \fo$ and $\eqso(\LL)$ is closed under subtraction by one, then $\shp$ is closed under subtraction by~one. 
\end{prop}
\proof
\input{proof-pi-minusone}
\qed
Therefore, the desired extension has to be achieved by allowing some local extensions to~$\Sigma_1$. More precisely, we define $\logex{1}$ as $\Sigma_1$ but allowing atomic formulae over a signature~$\R$ to be of the form either $u = v$ or $X(\bar u)$, where $X$ is a second-order variable, or $\varphi(\bar u)$, where $\varphi(\bar u)$ is a first-order formula over $\R$ (in particular, it does not mention any second-order variable). With this extension we obtain a class with the desired properties.
\begin{thm}\label{sigmafo-minusone}
The class $\eqso(\logex{1})$ is closed under sum, multiplication and subtraction by one. Moreover, $\eqso(\logex{1}) \subseteq \totp$ and every function in $\eqso(\logex{1})$ has an FPRAS.
\end{thm}
\proof
\input{proof-sigmafo-minusone}
\qed
The proof that $\eqso(\logex{1})$ is closed under subtraction by one is the most involved of the paper. We think the main technique used in this proof, which is based on considering some witnesses of logarithmic size, is of independent interest.

### 2018-12-09 LMCS/syntactic-horn.tex
all direct replacements by reviewer 1

In the previous section, we show that the class $\eqso(\logex{1})$ has good closure and approximation properties. Unfortunately, it is not clear whether it admits a {\em natural} complete problem under parsimonious reductions, where {\em natural} means any of the counting problems defined in this section or any other well-known counting problem (not one specifically designed to be complete for the class). Hence, in this section we follow a different approach to find a class of functions in $\shp$ with easy decision counterparts and natural complete problems, which is inspired by the approach followed in \cite{G92} that uses a restriction of second-order logic to Horn clauses for capturing $\ptime$ (over ordered structures). The following example shows how our approach works.
\begin{exa} \label{ex-hornsat-esop1}
Let $\R = \{\pP(\cdot,\cdot), \pN(\cdot,\cdot), \pV(\cdot), \pNC(\cdot),<\}$. This vocabulary is used as follows to encode a Horn formula. A fact $\pP(c,x)$ indicates that propositional variable $x$ is a disjunct in a clause $c$, while $\pN(c,x)$ indicates that $\neg x$ is a disjunct in $c$. Furthermore, $\pV(x)$ holds if  $x$ is a propositional variable, and $\pNC(c)$ holds if $c$ is a clause containing only negative literals, that is, $c$ is of the form $(\neg x_1 \vee \cdots \vee \neg x_n)$.

To define $\chsat$, we consider an \so-formula $\varphi(\pT)$ over $\R$, where $\pT$ is a unary predicate, such that for every Horn formula $\theta$ encoded by an $\R$-structure $\A$, the number of satisfying assignments of $\theta$ is equal to $\sem{\sa{\pT} \varphi(\pT)}(\A)$. In particular, $\pT(x)$ holds if, and only if, $x$ is a propositional variable that is assigned value true.  More specifically, 
%$\varphi(\pT)$ is defined as follows:
\begin{align*}
\varphi(\pT) \; :=\;\;  & \fa{x} (\pT(x) \to \pV(x)) \ \wedge\\
& \fa{c}  (\pNC(c) \to \ex{x} (\pN(c,x) \wedge \neg \pT(x))) \ \wedge\\
& \fa{c} \fa{x} ([\pP(c,x) \wedge \fa{y} (\pN(c,y) \to \pT(y))] \to \pT(x)).
\end{align*}
%Given that $\uhorn$ is designed with the goal in mind of capturing $\chsat$, we expect $\varphi(\pT)$ to be a formula in $\uhorn$. However, if we rewrite it as a conjunction of clauses we obtain the following:
We can rewrite $\varphi(\pT)$ in the following way:
\begin{align*}
& \fa{x}  (\neg \pT(x) \vee \pV(x)) \ \wedge\\
& \fa{c}  (\neg \pNC(c) \vee \ex{x} (\pN(c,x) \wedge \neg \pT(x)))\ \wedge\\
& \fa{c} \fa{x}  (\neg \pP(c,x) \vee \ex{y} (\pN(c,y) \wedge \neg \pT(y)) \vee \pT(x)).
\end{align*}
%The resulting formula $\varphi(\pT)$ is not in $\uhorn$, but it can be easily transformed into a formula in this class  by introducing an auxiliary binary predicate $\pA$ defined as follows:
Moreover, by introducing an auxiliary predicate $\pA$ defined as:
\begin{align*}
\fa{c} \fa{x}  (\neg \pA(c,x) \leftrightarrow [\pN(c,x) \wedge \neg \pT(x)]),
\end{align*}
we can translate $\varphi(\pT)$ into the following equivalent formula:
\begin{align*}
\psi(\pT,\pA) \; := \;\;  & \fa{x} (\neg \pT(x) \vee \pV(x)) \ \wedge\\
& \fa{c} (\neg \textit{NC}(c) \vee \ex{x} \neg \textit{A}(c,x)) \ \wedge\\
& \fa{c} \fa{x}  (\neg \textit{P}(c,x) \vee \ex{y} \neg \textit{A}(c,y) \vee \textit{T}(x)) \ \wedge\\
& \fa{c} \fa{x} (\neg \textit{N}(c,x) \vee \textit{T}(x) \vee \neg \textit{A}(c,x)) \ \wedge \\
& \fa{c} \fa{x} (\textit{A}(c,x) \vee \textit{N}(c,x)) \ \wedge\\
& \fa{c} \fa{x} (\textit{A}(c,x) \vee \neg\textit{T}(x)).
\end{align*}
More precisely, we have that:
\begin{align*}
\sem{\sa{\pT} \varphi(\pT)}(\A) &= \sem{\sa{\pT} \sa{\pA} \psi(\pT,\pA)}(\A),
\end{align*}
 for every $\R$-structure $\A$ encoding a Horn formula. Therefore, the formula $\psi(\pT,\pA)$ also defines $\chsat$. More importantly, $\psi(\pT,\pA)$ resembles a conjunction of Horn clauses except for the use of negative literals of the form $\ex{v} \neg \textit{A}(u,v)$.
\end{exa}
The previous example suggests that to define $\chsat$, we can use Horn formulae defined as follows. 
A positive literal is a formula of the form $X(\x)$, where $X$ is a second-order variable and $\x$ is a tuple of first-order variables, and a negative literal is a formula of the form $\ex{\v} \neg X(\u,\v)$, where $\u$ and $\v$ are tuples of first-order variables. Given a signature $\R$, a clause over $\R$ is a formula of the form $\fa{\x} (\varphi_1 \vee \cdots \vee \varphi_n)$, 
where each $\varphi_i$ ($1 \leq i \leq n$) is either a positive literal, a negative literal or an \fo-formula over $\R$.  A clause is said to be Horn if it contains at most one positive literal, and a formula is said to be Horn if it is a conjunction of Horn clauses. With this terminology, we define $\uhorn$ as the set of formulae $\psi$ such that $\psi$ is a Horn formula over a signature $\R$.

@@

\begin{prop}\label{prop:ehorn-pe}
$\eqso(\ehorn) \subseteq \totp$.
\end{prop}
For the sake of simplification, we postpone the proof the previous proposition after the proof of Theorem~\ref{sigma2hard}.

>>>>

In the previous section, we showed that the class $\eqso(\logex{1})$ has good closure and approximation properties. Unfortunately, it is not clear whether it admits a {\em natural} complete problem under parsimonious reductions, where {\em natural} means any of the counting problems defined in this section or any other well-known counting problem (not one specifically designed to be complete for the class). Hence, in this section we follow a different approach to find a class of functions in $\shp$ with easy decision counterparts and natural complete problems, which is inspired by the approach followed in \cite{G92} that uses a restriction of second-order logic to Horn clauses for capturing $\ptime$ (over ordered structures). The following example shows how our approach works.
\begin{exa} \label{ex-hornsat-esop1}
Let $\R = \{\pP(\cdot,\cdot), \pN(\cdot,\cdot), \pV(\cdot), \pNC(\cdot),<\}$. This vocabulary is used as follows to encode a Horn formula. A fact $\pP(c,x)$ indicates that propositional variable $x$ is a disjunct in a clause $c$, while $\pN(c,x)$ indicates that $\neg x$ is a disjunct in $c$. Furthermore, $\pV(x)$ holds if  $x$ is a propositional variable, and $\pNC(c)$ holds if $c$ is a clause containing only negative literals, that is, $c$ is of the form $(\neg x_1 \vee \cdots \vee \neg x_n)$.

To define $\chsat$, we consider an \so-formula $\varphi(\pT)$ over $\R$, where $\pT$ is a unary predicate, such that for every Horn formula $\theta$ encoded by an $\R$-structure $\A$, the number of satisfying assignments of $\theta$ is equal to $\sem{\sa{\pT} \varphi(\pT)}(\A)$. In particular, $\pT(x)$ holds if, and only if, $x$ is a propositional variable that is assigned value true.  More specifically, 
%$\varphi(\pT)$ is defined as follows:
\begin{align*}
\varphi(\pT) \; :=\;\;  & \fa{x} (\pT(x) \to \pV(x)) \ \wedge\\
& \fa{c}  (\pNC(c) \to \ex{x} (\pN(c,x) \wedge \neg \pT(x))) \ \wedge\\
& \fa{c} \fa{x} ([\pP(c,x) \wedge \fa{y} (\pN(c,y) \to \pT(y))] \to \pT(x)).
\end{align*}
%Given that $\uhorn$ is designed with the goal in mind of capturing $\chsat$, we expect $\varphi(\pT)$ to be a formula in $\uhorn$. However, if we rewrite it as a conjunction of clauses we obtain the following:
We can rewrite $\varphi(\pT)$ in the following way:
\begin{align*}
& \fa{x}  (\neg \pT(x) \vee \pV(x)) \ \wedge\\
& \fa{c}  (\neg \pNC(c) \vee \ex{x} (\pN(c,x) \wedge \neg \pT(x)))\ \wedge\\
& \fa{c} \fa{x}  (\neg \pP(c,x) \vee \ex{y} (\pN(c,y) \wedge \neg \pT(y)) \vee \pT(x)).
\end{align*}
%The resulting formula $\varphi(\pT)$ is not in $\uhorn$, but it can be easily transformed into a formula in this class  by introducing an auxiliary binary predicate $\pA$ defined as follows:
Moreover, by introducing an auxiliary predicate $\pA$ defined as:
\begin{align*}
\fa{c} \fa{x}  (\neg \pA(c,x) \leftrightarrow [\pN(c,x) \wedge \neg \pT(x)]),
\end{align*}
we can translate $\varphi(\pT)$ into the following equivalent formula:
\begin{align*}
\psi(\pT,\pA) \; := \;\;  & \fa{x} (\neg \pT(x) \vee \pV(x)) \ \wedge\\
& \fa{c} (\neg \textit{NC}(c) \vee \ex{x} \neg \textit{A}(c,x)) \ \wedge\\
& \fa{c} \fa{x}  (\neg \textit{P}(c,x) \vee \ex{y} \neg \textit{A}(c,y) \vee \textit{T}(x)) \ \wedge\\
& \fa{c} \fa{x} (\neg \textit{N}(c,x) \vee \textit{T}(x) \vee \neg \textit{A}(c,x)) \ \wedge \\
& \fa{c} \fa{x} (\textit{A}(c,x) \vee \textit{N}(c,x)) \ \wedge\\
& \fa{c} \fa{x} (\textit{A}(c,x) \vee \neg\textit{T}(x)).
\end{align*}
More precisely, we have that:
\begin{align*}
\sem{\sa{\pT} \varphi(\pT)}(\A) &= \sem{\sa{\pT} \sa{\pA} \psi(\pT,\pA)}(\A),
\end{align*}
 for every $\R$-structure $\A$ encoding a Horn formula. Therefore, the formula $\psi(\pT,\pA)$ also defines $\chsat$. More importantly, $\psi(\pT,\pA)$ resembles a conjunction of Horn clauses except for the use of negative literals of the form $\ex{v} \neg \textit{A}(u,v)$. \qed
\end{exa}
The previous example suggests that to define $\chsat$, we can use Horn formulae defined as follows. 
A positive literal is a formula of the form $X(\x)$, where $X$ is a second-order variable and $\x$ is a tuple of first-order variables, and a negative literal is a formula of the form $\ex{\v} \neg X(\u,\v)$, where $\u$ and $\v$ are tuples of first-order variables. Given a signature $\R$, a clause over $\R$ is a formula of the form $\fa{\x} (\varphi_1 \vee \cdots \vee \varphi_n)$, 
where each $\varphi_i$ ($1 \leq i \leq n$) is either a positive literal, a negative literal or an \fo-formula over $\R$.  A clause is said to be Horn if it contains at most one positive literal, and a formula is said to be Horn if it is a conjunction of Horn clauses. With this terminology, we define $\uhorn$ as the set of formulae $\psi$ such that $\psi$ is a Horn formula over a signature $\R$.

@@

\begin{prop}\label{prop:ehorn-pe}
$\eqso(\ehorn) \subseteq \totp$.
\end{prop}
For the sake of simplicity, we postpone the proof of the previous proposition to after the proof of Theorem~\ref{sigma2hard}.

### 2018-12-09 LMCS/syntactic.tex
all direct replacements by reviewer 1

>>>>

The class $\shp$ was introduced in \cite{Valiant79} to prove that computing the permanent of a matrix, as defined in Example \ref{exa-perm}, is a $\shp$-complete problem.
As a consequence of this result many counting problems have been proved to be $\shp$-complete~\cite{V79b,arora2009computational}.
%\cite{V79b,PB83,P86,L86,BW91,HMRS98,DG00,BW05,DS12, PS13,PS14}.
Among them, problems having easy decision counterparts play a fundamental role, as a counting problem with a hard decision version is expected to be hard. Formally, the decision problem associated to a function $f : \Sigma^* \to \mathbb{N}$ is defined as $L_f = \{ x \in \Sigma^* \mid f(x) > 0 \}$, and $f$ is said to have an easy decision version if $L_f \in \ptime$. 
Many prominent examples satisfy this property, like computing the number of: perfect matchings of a bipartite graph ($\cpm$)~\cite{Valiant79}, satisfying assignments of a DNF ($\cdnf$)~\cite{DHK05,KL83} or Horn ($\chsat$)~\cite{V79b} propositional formula, among others.

Counting problems with easy decision versions play a fundamental role in the search for efficient approximation algorithms for functions in $\shp$. A fully-polynomial randomized approximation scheme (FPRAS) for a function $f : \Sigma^* \to \bbN$ is a randomized algorithm $\mathcal{A} : \Sigma^* \times (0,1) \to \bbN$ such that: (1) for every string $x \in \Sigma^*$ and real value $\varepsilon \in (0,1)$, the probability that $|f(x) - \mathcal{A}(x,\varepsilon)| \leq \varepsilon \cdot f(x)$ is at least $\frac{3}{4}$, and (2) the running time of $\mathcal{A}$ is polynomial in the size of $x$ and $1/\varepsilon$ \cite{KL83}. Notably, there exist $\shp$-complete functions that can be efficiently approximated as they admit FPRAS; for instance, there exist FPRAS for $\cdnf$~\cite{KL83} and $\cpm$~\cite{JSV04}. A key observation here is that if a function $f$ admits an FPRAS, then $L_f$ is in the randomized complexity class $\bpp$~\cite{G77}. Hence, under the widely believed assumption that  $\np \not\subseteq \bpp$, we cannot hope for an FPRAS for a function in $\shp$ whose decision counterpart is $\np$-complete, and we have to concentrate on the class of counting problems with easy decision versions.

The importance of the class of counting problems with easy decision counterparts has motivated the search for robust classes of functions in $\shp$ with this property \cite{PagourtzisZ06}. But the key question here is what should be considered a {\em robust} class. 
A first desirable condition has to do with the closure properties satisfied by the class, which is a common theme when studying function complexity classes \cite{OH93,FH08}. Analogously to the cases of $\ptime$ and $\np$, which are closed under intersection and union, we expect our class to be closed under multiplication and sum. For a more elaborated closure property, assume that $\textit{sat\_one}$ is a function that returns one plus the number of satisfying assignments of a propositional formula. Clearly $\textit{sat\_one}$ is a $\shp$-complete function whose decision counterpart $L_{\textit{sat\_one}}$ is trivial. But should $\textit{sat\_one}$ be part of a robust class of counting functions with easy decision versions? The key insight here is that if a function in $\shp$ has an easy decision counterpart $L$, then as $L \in \np$ we expect  to have a polynomial-time algorithm that verifies whether $x \in L$ by constructing witnesses for~$x$. 
%As an example of this considered all the counting functions mentioned in the previous paragraphs and, in particular, the decision counterpart of $\cdnf$ that can be solved by explicitly constructing in polynomial-time a satisfying assignment for an input propositional formula. 
Moreover, if such an algorithm for constructing witnesses exists, then we also expect to be able to manipulate such witnesses and in some cases to remove them. In other words, we expect a robust class $\CC$ of counting functions with easy decision versions to be closed under subtraction by one, that is, if $g \in \CC$, then the function $g \dotdiv 1$ should also be in $\CC$, where $(g \dotdiv 1)(x)$ is defined as $g(x) - 1$ if $g(x) \geq 1$, and as~$0$ otherwise. Notice that, unless $\ptime = \np$, no such class can contain the function $\textit{sat\_one}$ because $\textit{sat\_one} \dotdiv 1$ counts the number of satisfying assignments of a propositional formula.

A second desirable condition of robustness is the existence of natural complete problems~\cite{P94}. Special attention has to be paid here to the notion of reduction used for completeness. Notice that under the notion of Cook reduction, originally used in \cite{Valiant79}, the problems $\cdnf$ and $\csat$ are $\shp$-complete. However, $\cdnf$ has an easy decision counterpart and admits an FPRAS, while $\csat$ does not satisfy these conditions unless $\ptime = \np$. Hence a more strict notion of reduction has to be considered; in particular, the notion of parsimonious reduction (to be defined later) satisfies that if a function $f$ is parsimoniously reducible to a function $g$, then $L_g \in \ptime$ implies that $L_f \in \ptime$ and the existence of an FPRAS for $g$ implies the existence of a FPRAS for $f$.

In this section, we use the framework developed in this paper to address the problem of defining a robust class of functions with easy decision versions. More specifically, we use the framework to introduce in Section \ref{sec-hier-shp} a syntactic hierarchy of counting complexity classes contained in $\shp$. Then this hierarchy is used in Section \ref{sec-clo} to define a class of functions with easy decision versions and good closure properties, and in Section \ref{sec-horn} to define a class of functions with easy decision versions and 
natural complete problems.
